\documentclass[conference]{IEEEtran}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\hypersetup{hidelinks}

\title{A Graph-Based Mixed-Integer Formulation for Spiral Galaxies and Center-Guided Region Partitioning}
\author{
\IEEEauthorblockN{Santiago Florido Gomez}
\IEEEauthorblockA{\textit{RO203, Dipl\^ome d'ing\'enieur}\\
\textit{ENSTA Paris}\\
Palaiseau, France\\
santiago.florido@ensta.fr}
\and
\IEEEauthorblockN{Jose Daniel Chacon Gomez}
\IEEEauthorblockA{\textit{RO203, Dipl\^ome d'ing\'enieur}\\
\textit{ENSTA Paris}\\
Palaiseau, France\\
jose-daniel.chacon@ensta.fr}
}

\begin{document}
\pagenumbering{arabic}
\pagestyle{plain}
\maketitle
\thispagestyle{plain}

\begin{abstract}
This paper presents the development of a graph-based exact formulation for the
\textit{Spiral Galaxies} puzzle and studies its relevance as a model for
center-guided connected region partitioning. The board is represented as a grid
graph in which each feasible galaxy is defined as a connected region assigned
to a unique center and constrained by admissibility and central symmetry. The
resulting problem is formulated as a mixed-integer linear model combining
binary assignment variables with flow-based constraints to enforce
connectivity. Beyond the mathematical model, the project includes a Python
implementation of two complementary solver backends, validation tools for
feasibility checks, a fixed-range dataset-generation pipeline, an empirical
hard-threshold search utility driven by the exact solver, and a Pygame
interface for interactive visualization and experimentation. The article
describes the modeling choices, the software architecture used to translate the
formulation into code, and the role of this framework as a compact case study
in exact discrete optimization with geometric priors.
\end{abstract}

\begin{IEEEkeywords}
graph theory, constraint solving, puzzle modeling, pygame
\end{IEEEkeywords}

\input{sections/01_introduction.tex}
\input{sections/02_mathematical_model.tex}
\input{sections/03_methodology.tex}
\input{sections/04_results.tex}
\input{sections/05_conclusion.tex}

\begin{thebibliography}{00}
\bibitem{fertin2015}
G. Fertin, S. Jamshidi, and C. Komusiewicz, ``Towards an algorithmic guide to Spiral Galaxies,'' \emph{Theoretical Computer Science}, vol. 586, pp. 26--39, 2015, doi: 10.1016/j.tcs.2015.01.051.

\bibitem{shi2000}
J. Shi and J. Malik, ``Normalized Cuts and Image Segmentation,'' \emph{IEEE Transactions on Pattern Analysis and Machine Intelligence}, vol. 22, no. 8, pp. 888--905, 2000, doi: 10.1109/34.868688.

\bibitem{validi2022}
H. Validi, A. Buchanan, and E. Lykhovyd, ``Imposing Contiguity Constraints in Political Districting Models,'' \emph{Operations Research}, vol. 70, no. 2, pp. 867--892, 2022, doi: 10.1287/opre.2021.2141.

\bibitem{miyazawa2021}
F. K. Miyazawa, P. F. S. Moura, M. J. Ota, and Y. Wakabayashi, ``Partitioning a graph into balanced connected classes: Formulations, separation and experiments,'' \emph{European Journal of Operational Research}, vol. 293, no. 3, pp. 826--836, 2021, doi: 10.1016/j.ejor.2020.12.059.

\bibitem{sun2012}
Y. Sun and B. Bhanu, ``Reflection symmetry-integrated image segmentation,'' \emph{IEEE Transactions on Pattern Analysis and Machine Intelligence}, vol. 34, no. 9, pp. 1827--1841, 2012, doi: 10.1109/TPAMI.2011.259.

\bibitem{gorelick2017}
L. Gorelick, O. Veksler, Y. Boykov, and C. Nieuwenhuis, ``Convexity Shape Prior for Binary Segmentation,'' \emph{IEEE Transactions on Pattern Analysis and Machine Intelligence}, vol. 39, no. 2, pp. 258--271, 2017, doi: 10.1109/TPAMI.2016.2547399.

\bibitem{dey2024}
N. Dey, S. M. Abulnaga, B. Billot, E. A. Turk, P. E. Grant, A. V. Dalca, and P. Golland, ``AnyStar: Domain Randomized Universal Star-Convex 3D Instance Segmentation,'' in \emph{Proceedings of the IEEE/CVF Winter Conference on Applications of Computer Vision (WACV)}, 2024, doi: 10.1109/WACV57701.2024.00742.
\end{thebibliography}

\end{document}
