\section{Results}

The experimental dataset was generated with the fixed-range pipeline described
in the implementation section, rather than through the threshold-search
utility. More precisely, the command-line generator produced square instances
for all three difficulty levels on the range from \(7\times 7\) to
\(11\times 11\), with \(20\) valid instances retained for each board size and
each difficulty label. For every candidate, the pipeline executed profile-aware
center placement, constructive region growth, partition closure, optional shape
perturbation, and final certification with the exact flow backend. Only
instances whose certified assignment passed full structural validation were
stored. Since these board sizes extend the canonical profile sizes, the center
count bounds were rescaled as a function of board area while preserving the
qualitative targets attached to each difficulty profile, namely the
cell/edge/vertex center mix, the admissible-domain overlap range, the
irregularity target, and the minimum number of non-rectangular regions. The
dataset was generated on square boards from \(7\times 7\) to \(11\times 11\),
with \(20\) retained instances per size and difficulty, \(64\) generation
retries per seeded attempt, \(24\) outer seed sweeps, \(16\) seed blocks, and
exact-flow screening/certification under the global public time limit of
\(0.5\) seconds per solver call.

\begin{table}[t]
\caption{Dataset-generation parameters used in the experimental campaign.}
\label{tab:dataset_generation_parameters}
\centering
\scriptsize
\setlength{\tabcolsep}{2.6pt}
\renewcommand{\arraystretch}{1.05}
\resizebox{\columnwidth}{!}{%
\begin{tabular}{lccc}
\toprule
Parameter & E & M & H \\
\midrule
Canonical center bounds & \([2,5]\) & \([6,10]\) & \([10,16]\) \\
Center mix (c/e/v) & \(0.70/0.25/0.05\) & \(0.45/0.40/0.15\) & \(0.25/0.45/0.30\) \\
Overlap target & \([0.00,0.25]\) & \([0.20,0.45]\) & \([0.35,0.65]\) \\
Irregularity target & \([0.00,0.10]\) & \([0.08,0.22]\) & \([0.20,0.45]\) \\
Min. non-rect. regions & \(0\) & \(1\) & \(2\) \\
Uniqueness flag & true & true & true \\
Target area \(a_d\) & \(16\) & \(11\) & \(8\) \\
Variation \(\delta_d\) & \(1\) & \(1\) & \(2\) \\
Feasibility cap \(r_d\) & \(4\) & \(3\) & \(2\) \\
\bottomrule
\end{tabular}
}
\end{table}

The size-aware scaling rule used to derive the center-count interval from the
board area is described in the dataset-construction subsection of the
methodology. Table~\ref{tab:dataset_generation_parameters} summarizes the
numerical parameters that were used in the experimental campaign.
